Bricking up and overgrowing backdoors in Intrusion Detection Systems

### 2019-09-08 ex2/paper/paper.tex
Improved pruning_example figure

\begin{table*}[b]
\caption{Detection performance results.} \label{tab:performance_results}
\begin{tabular}{l r r r r} \toprule
& \multicolumn{2}{r}{UNSW-NB15} & \multicolumn{2}{r}{CIC-IDS-2017} \\
& RF & DL & RF & DL \\ \midrule
Accuracy	& 0.990 $\pm$ 0.000	& 0.989 $\pm$ 0.000	& 0.997 $\pm$ 0.000	& 0.998 $\pm$ 0.000	\\
Precision	& 0.854 $\pm$ 0.001	& 0.845 $\pm$ 0.004	& 0.997 $\pm$ 0.000	& 0.999 $\pm$ 0.000	\\
Recall	& 0.850 $\pm$ 0.003	& 0.829 $\pm$ 0.009	& 0.993 $\pm$ 0.000	& 0.992 $\pm$ 0.000	\\
F1 score	& 0.852 $\pm$ 0.002	& 0.837 $\pm$ 0.003	& 0.995 $\pm$ 0.000	& 0.995 $\pm$ 0.000	\\
Youden's J	& 0.845 $\pm$ 0.003	& 0.823 $\pm$ 0.009	& 0.992 $\pm$ 0.000	& 0.991 $\pm$ 0.000	\\
Backdoor acc.	& 1.000 $\pm$ 0.000	& 0.998 $\pm$ 0.002	& 1.000 $\pm$ 0.000	& 1.000 $\pm$ 0.000	\\

@@

@@

We thus propose a defense approach that can be used for similar use cases like the mechanisms recently proposed to remove backdoors from \glspl{cnn}. The concept behind this pruning defense is that leaves that are never used by samples in the validation dataset might be used by backdoored inputs. If these ``useless'' leaves are removed, performance of the classifier on the validation dataset should not decrease by a lot while decisions that are used for the backdoor are hopefully removed. We develop several variants of our pruning defense with increasing levels of sophistication. These iterations are pruning
\begin{enumerate}
\item leaves based on their usage: This means that the least used leaf is pruned first and the most used one last. 
\item only ``harmless'' leaves based on their usage: In the case of \glspl{nids} we assume that attackers want to make attacks appear to be benign. 
\item only ``harmless'' leaves based on their usage and based on their depth if two leaves have an equal number of usages: The rationale behind this is that we assume that hand-crafted backdoors require fewer rules for a decision tree to classify than regular samples. Thus, samples higher up the tree (with lower depth) are pruned first, if they have the same number of usages like another leaf with a higher depth. 
\end{enumerate}
\begin{figure*}[h]
\includegraphics[width=\textwidth]{pruning_example.pdf}
\caption{Example of a decision tree being pruned. To the left of each leaf is \textcolor{MidnightBlue}{the number of times a leaf was used} by validation samples and its \textcolor{Fuchsia}{depth} in the decision tree is to the right of each leaf. It is also indicated that a sample is \protect\sethlcolor{green} \hl{harmless} (leaves with a 0 inside) because these are the ones our algorithm considers as candidates for pruning. The leaf that is going to be cut in each step has a dashed connection to its parent. This decision tree is not functional anymore in the end because all its leaves indicate that a sample is an attack. However, the decision trees that we prune usually have several thousands of leaves and we never prune until the very end and thus this case of a degenerate decision tree is not problematic.}
\label{fig:pruningExample}
\end{figure*}
\autoref{fig:pruningExample} shows an example of the last iteration of pruning applied to a tree.

@@

\begin{figure*}[p]
\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/sourceTransportPort_nn.pdf}
\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/sourceTransportPort_rf.pdf}

@@

We generated PD plots for all features contained in our extracted feature vector. Fig.~\ref{fig:pdp} shows examples for some features for the different folds. Comparing plots generated for our deep learning and for our random forest approach, respectively, the two machine learning approaches show the difference that deep learning PDPs are smooth, while random forest PDPs fluctuate heavily. Considering the functioning of neural networks and decision trees, this is an expected result.

Considering the explanations the PD plots provide, consistency between deep learning and random forest can be observed for some features, while for other features both methods provide substantially different explanations. For example, the two upper plots in Fig.~\ref{fig:pdp} exhibit a very similar behavior, while the lower plots show remarkable differences between both methods. An additional observation is, however, that random forests show substantial differences between the folds.

@@

The topmost plots in Fig.~\ref{fig:pdp} show the PDP for source ports, which intuitively provides important information for attack detection. Indeed, the plots show that predictions vary substantially with this feature. Furthermore, attack traffic mainly uses higher (unprivileged) ports.

@@

Most importantly, however, the \textit{mean\_srcTTL} feature depicted in Fig.~\ref{fig:pdp} shows that both deep learning and random forest learn to consider very specific TTL values for their decisions. This points out that attack traffic was generated from a very specific network with specific TTL values throughout dataset generation.
By no means can this behaviour generalize to network traffic in real deployments.

@@

\begin{figure*}[p]
\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/sourceTransportPort_nn.pdf}
\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/sourceTransportPort_rf.pdf}

@@

In addition to PD plots, we generated ALE plots for all features in our feature vectors. Fig.~\ref{fig:ale} depicts plots for the same features as presented in Fig.~\ref{fig:pdp}.

As described in Section~\ref{sec:plots}, ALE plots might provide benefits in explaining the models' decisions and, hence, seem particularly important.

Plots depicted in Fig.~\ref{fig:ale} show a certain similarity to the PD plots in Fig.\ref{fig:pdp}. However, as apparant at the \textit{srcPort}, explanations from PD and ALE plots might also deviate substantially \todo{Are we sure that this is not a bug?}.

@@

% AH This figures currently serve as placeholder/example. TOOD: decide what to plot exactly
\begin{figure}[t]
\includegraphics[width=\columnwidth]{../prune_CAIA_backdoor_15/prune.pdf}
\caption{Performance metrics throughout the pruning process.}
\end{figure}

\begin{figure}[t]
\includegraphics[width=\columnwidth]{../prune_CAIA_backdoor_17/{prune_1.00_nn_0_bd}.pdf}
\caption{Correlation coefficient of neuron activation with backdoor usage throughout the pruning process.}
\end{figure}

\begin{figure*}[p]
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/sourceTransportPort_nn_bd.pdf}
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/sourceTransportPort_rf_bd.pdf}

@@

%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTotalLength),forward)_nn_bd.pdf}
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTotalLength),forward)_rf_bd.pdf}

@@

\begin{table}
\caption{Detection performance results for models with backdoor.}
\label{tab:performance_results_bd}
\begin{tabular}{l r r} \toprule
& Random Forest & Deep Learning \\ \midrule
Accuracy	&	0.9986 $\pm$ 0.0004	&	0.9974 $\pm$ 0.0001		\\
Precision	&	0.9981 $\pm$ 0.0005	&	0.9986 $\pm$ 0.0004		\\
Recall	&	0.9964 $\pm$ 0.0009	&	0.9912 $\pm$ 0.0004		\\
F1	&	0.9972 $\pm$ 0.0007	&	0.9949 $\pm$ 0.0002		\\
Youden	&	0.9957 $\pm$ 0.0011	&	0.9907 $\pm$ 0.0003		\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{Efficacy of the backdoor.}
\begin{tabular}{l r r} \toprule
 & Random Forest & Deep Learning \\ \midrule
Forward & 1 & 0,9999 $\pm$ 0,0001 \\
Backward & 1 & 0,9999 $\pm$ 0,0000 \\
\bottomrule
\end{tabular}
\end{table}

\end{document}
\endinput
%%
%% End of file `sample-sigconf.tex'.

>>>>

\begin{table*}[h]
\caption{Detection performance results.} \label{tab:performance_results}
\begin{tabular}{l r r r r} \toprule
& \multicolumn{2}{r}{UNSW-NB15} & \multicolumn{2}{r}{CIC-IDS-2017} \\
& RF & DL & RF & DL \\ \midrule
Accuracy	& 0.990 $\pm$ 0.000	& 0.989 $\pm$ 0.000	& 0.997 $\pm$ 0.000	& 0.998 $\pm$ 0.000	\\
Precision	& 0.854 $\pm$ 0.001	& 0.845 $\pm$ 0.004	& 0.997 $\pm$ 0.000	& 0.999 $\pm$ 0.000	\\
Recall	& 0.850 $\pm$ 0.003	& 0.829 $\pm$ 0.009	& 0.993 $\pm$ 0.000	& 0.992 $\pm$ 0.000	\\
F1 score	& 0.852 $\pm$ 0.002	& 0.837 $\pm$ 0.003	& 0.995 $\pm$ 0.000	& 0.995 $\pm$ 0.000	\\
Youden's J	& 0.845 $\pm$ 0.003	& 0.823 $\pm$ 0.009	& 0.992 $\pm$ 0.000	& 0.991 $\pm$ 0.000	\\
Backdoor acc.	& 1.000 $\pm$ 0.000	& 0.998 $\pm$ 0.002	& 1.000 $\pm$ 0.000	& 1.000 $\pm$ 0.000	\\

@@

\begin{figure}[h]
\includegraphics[width=\columnwidth]{../prune_CAIA_backdoor_17/{prune_1.00_nn_0_bd}.pdf}
\caption{Correlation coefficient of neuron activation with backdoor usage throughout the pruning process.}
\end{figure}

@@

We thus propose a defense approach that can be used for similar use cases like the mechanisms recently proposed to remove backdoors from \glspl{cnn}. The concept behind this pruning defense is that leaves that are never used by samples in the validation dataset might be used by backdoored inputs. If these ``useless'' leaves are removed, performance of the classifier on the validation dataset should not decrease by a lot while decisions that are used for the backdoor are hopefully removed. We develop several variants of our pruning defense with increasing levels of sophistication. These iterations are pruning
\begin{enumerate}
\item leaves based on their usage: This means that the least used leaf is pruned first and the most used one last. 
\item only ``harmless'' leaves based on their usage: In the case of \glspl{nids} we assume that attackers want to make attacks appear to be benign. 
\item only ``harmless'' leaves based on their usage and based on their depth if two leaves have an equal number of usages: The rationale behind this is that we assume that hand-crafted backdoors require fewer rules for a decision tree to classify than regular samples. Thus, samples higher up the tree (with lower depth) are pruned first, if they have the same number of usages like another leaf with a higher depth. 
\end{enumerate}
\begin{figure*}[h]
\includegraphics[width=\textwidth]{pruning_example.pdf}
\caption{Example of a decision tree being pruned. To the left of each leaf is \textcolor{MidnightBlue}{the number of times a leaf was used} by validation samples and its \textcolor{Fuchsia}{depth} in the decision tree is to the right of each leaf. It is also indicated that a sample is \protect\sethlcolor{green} \hl{harmless} (leaves with a 0 inside) because these are the ones our algorithm considers as candidates for pruning. The leaf that is going to be cut in each step has a dashed connection to its parent. After step 2 the algorithm stops as pruning the last remaining leaf would make the decision tree non-functional.}
\label{fig:pruningExample}
\end{figure*}
\autoref{fig:pruningExample} shows an example of the last iteration of pruning applied to a tree.

\begin{figure}[h]
\includegraphics[width=\columnwidth]{../prune_CAIA_backdoor_15/prune.pdf}
\caption{Performance metrics throughout the pruning process. }
\label{fig:performancePruningRF}
\end{figure}

\autoref{fig:performancePruningRF} shows the for the UNSW-NB15 dataset pruning not only

@@

\begin{figure*}[h]
\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/sourceTransportPort_nn.pdf}
\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/sourceTransportPort_rf.pdf}

@@

We generated PD plots for all features contained in our extracted feature vector. \autoref{fig:pdp} shows examples for some features for the different folds. Comparing plots generated for our deep learning and for our random forest approach, respectively, the two machine learning approaches show the difference that deep learning PDPs are smooth, while random forest PDPs fluctuate heavily. Considering the functioning of neural networks and decision trees, this is an expected result.

Considering the explanations the PD plots provide, consistency between deep learning and random forest can be observed for some features, while for other features both methods provide substantially different explanations. For example, the two upper plots in \autoref{fig:pdp} exhibit a very similar behavior, while the lower plots show remarkable differences between both methods. An additional observation is, however, that random forests show substantial differences between the folds.

@@

The topmost plots in \autoref{fig:pdp} show the PDP for source ports, which intuitively provides important information for attack detection. Indeed, the plots show that predictions vary substantially with this feature. Furthermore, attack traffic mainly uses higher (unprivileged) ports.

@@

Most importantly, however, the \textit{mean\_srcTTL} feature depicted in \autoref{fig:pdp} shows that both deep learning and random forest learn to consider very specific TTL values for their decisions. This points out that attack traffic was generated from a very specific network with specific TTL values throughout dataset generation.
By no means can this behaviour generalize to network traffic in real deployments.

@@

% EVIL FIGURE: It suppresses the two "detection" figures
\begin{figure*}[h]
\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/sourceTransportPort_nn.pdf}
\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/sourceTransportPort_rf.pdf}

@@

In addition to PD plots, we generated ALE plots for all features in our feature vectors. \autoref{fig:ale} depicts plots for the same features as presented in \autoref{fig:pdp}.

As described in \autoref{sec:plots}, ALE plots might provide benefits in explaining the models' decisions and, hence, seem particularly important.

Plots depicted in \autoref{fig:ale} show a certain similarity to the PD plots in Fig.\ref{fig:pdp}. However, as apparant at the \textit{srcPort}, explanations from PD and ALE plots might also deviate substantially \todo{Are we sure that this is not a bug?}.

@@

\begin{figure*}[h]

@@

@@

\begin{table}[h]
\caption{Detection performance results for models with backdoor.}
\label{tab:performance_results_bd}
\begin{tabular}{l r r} \toprule
& Random Forest & Deep Learning \\ \midrule
Accuracy	&	0.9986 $\pm$ 0.0004	&	0.9974 $\pm$ 0.0001		\\
Precision	&	0.9981 $\pm$ 0.0005	&	0.9986 $\pm$ 0.0004		\\
Recall	&	0.9964 $\pm$ 0.0009	&	0.9912 $\pm$ 0.0004		\\
F1	&	0.9972 $\pm$ 0.0007	&	0.9949 $\pm$ 0.0002		\\
Youden	&	0.9957 $\pm$ 0.0011	&	0.9907 $\pm$ 0.0003		\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{Efficacy of the backdoor.}
\begin{tabular}{l r r} \toprule
 & Random Forest & Deep Learning \\ \midrule
Forward & 1 & 0,9999 $\pm$ 0,0001 \\
Backward & 1 & 0,9999 $\pm$ 0,0000 \\
\bottomrule
\end{tabular}
\end{table}

\end{document}
